The Disk Wind Contribution to the Optical Spectra of Cataclysmic Variables

### 2014-02-10 paper.tex
Figures subdirectory
Some changes, new figures

%%%% Various options for document class.
\documentclass[usenatbib, a4paper]{mn2e}
%%\documentclass[preprint2, a4paper, 11pt]{aastex}
%%\documentclass[twocolum]{revtex4}
%%\documentclass{report}
% \documentclass[useAMS,usenatbib,referee]{mn2e}

@@

\title{The Disk Wind Contribution to Optical Lines in Cataclysmic Variables}
\author{J. H.
  Matthews, C. Knigge, K. S. Long, S. A. Sim \& N. Higginbottom}
%%\altaffiltext{1}{School of Physics \& Astronomy, University of Southampton,
  %%Southampton, SO17 1BJ, UK}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          ABSTRACT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle

\begin{abstract}
Disk-dominated High-State Cataclysmic Variable systems (CVs) exhibit strong emission in optical recombination lines.
Here we present results obtained by improving a Monte Carlo radiative transfer code originally used
to model UV resonance lines in CVs, which show the potential contribution to these optical lines
from a line driven accretion disk wind.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          INTRODUCTION
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Cataclysmic Variables (CVs), are systems in which a white dwarf accretes matter from a donor star
via Roche-lobe overflow. In Disk-dominated CVs, this accretion is mediated by an accretion disk which forms 
around the white dwarf, and it has been shown that the spectra of high state CVs
exhibit absorption features believed to be caused by an outflow, potentially line-drive,
rising from the accretion disk.

The ultraviolet spectra of CVs, in particular, exhibity P-Cygni like profiles in resonance lines such as 
NV, OVI and, most commonly, CIV.

@@

\py is a Monte Carlo radiative transfer code which uses the Sobolev approximation. It is originally
introduced by Long \& Knigge (2002), and a more up-to-date description is provided by
Higginbottom et al. (2013).

\subsection{Macro-atoms}
In its standard operating mode, \py uses a two-level atom approximation (see e.g Mihalas 1982\nocite{mihalas}, Ch. 11) for its treatment of lines. This approximation works well when treating so-called `resonance lines' (such as \civ, O\textsc{vi}), in which the excited electron is strongly coupled to the ground state, but breaks down for more complex situations such as recombination cascades. To properly model recombination lines, including the Lyman and Balmer series, a more complete treatment of line transfer is required.

@@

@@

\section{Testing the macro atom scheme}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{caseA.png}
\caption{Case A comparison between S59 and \py at 10000K and 20000K}
\end{figure}

\begin{figure}
\includegraphics[width=0.5\textwidth]{caseB.png}
\caption{Case B comparison between S59 and \py at 10000K and 20000K}
\end{figure}

@@

\section{The CV Model}
As in LK02, we follow the prescription of SV93 in both our fundamental biconical wind model and velocity law

@@

\section{Results from Simulations}

@@

\section{Discussion}

@@

>>>>

%%%% Various options for document class.
%\documentclass[usenatbib, a4paper]{preprint}
%\documentclass[preprint, a4paper, 11pt]{aastex}
%%\documentclass[twocolum]{revtex4}
%%\documentclass{report}
\documentclass[useAMS,usenatbib, onecolumn]{mn2ejm}

@@

\title{The Disk Wind Contribution to the Optical Spectra of Cataclysmic Variables}
\author{J. H.
  Matthews, C. Knigge, K. S. Long, S. A. Sim \& N. Higginbottom}
%%\altaffiltext{1}{School of Physics \& Astronomy, University of Southampton,
  %%Southampton, SO17 1BJ, UK}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          ABSTRACT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\maketitle

\begin{abstract}
High-State non-magnetic Cataclysmic Variable systems (CVs) exhibit strong emission in Hydrogen \& Helium 
optical recombination lines. Here we present results obtained by encorporating a macro atom treatment into
our Monte Carlo radiative transfer code originally used
to model UV resonance lines in CVs, \py. Our benchmark CV model is capable of producing
all of the notable optical lines (e.g. He II 4686, H${\alpha}$, He I), as well as
enchanced emission in the Ly-$\alpha$ and He II $1640$ UV lines.
In addition, the improved treatment of recombination means that the bound-free continuum
emission is sufficient to mask the so-called `Balmer jump' photoabsorption edge, 
suggesting a potential solution to a longstanding problem. 
\end{abstract}

%%\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          INTRODUCTION
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Non-magnetic Cataclysmic Variables (CVs), are systems in which a white dwarf accretes matter from a donor star
through Roche-lobe overflow onto an accretion disk. It has been shown that the spectra of high state CVs
exhibit absorption features believed to be caused by an outflow, potentially line-driven,
rising from the accretion disk \citep{cordova1982}. 
The ultraviolet spectra of CVs, in particular, exhibit P-Cygni like profiles in resonance lines such as 
NV, OVI and, most commonly, CIV. This hints at the possibility of the wind being line driven, 
as these resonance lines must experience a line force if they are photoionised by a disk or other photon source
close to the centre of the system.

Multiple attempts have been made to model UV wind spectra by treating the wind
as a biconical flow arising from the accretion disk, and conducting radiative transfer simulations to attempt
to reproduce the resonance lines that appear in the spectra. The first generation of models
applied knowledge gleaned from stellar wind modeling to the problem (Mauche \& Raymond
1987; Drew 1987; Vitello \& Shlosman 1988), showing that mass loss rates of $\sim10\%$ of the accretion rate
 were required to produce the required line strengths.
Shlosman \& Vitello (1993; hereafter SV93) describe a radiative transfer code which first computes
an ionization structure based on constant wind temperature, and then used the Sobolev approximation
for line transfer.

%% reorganise, mention SW Sex, UX UMa systems?

@@

\py is a Monte Carlo radiative transfer code which uses the Sobolev approximation.  
It expands on the work of SV93 and also incorporates techniques described by
Lucy (1999a, 1999b, 2002, 2003). The code is first described by Long \& Knigge (2002),
who synthesized the UV spectra of CVs, successfully reproducing P-Cygni resonance line profiles
whilst solving self-consistently for the thermal balance and ionization structure in the wind
according to the method of Mazzali \& Lucy (1993). \py has since been used with application
to QSO wind modeling \citep{higginbottom2013}, which involved a new ionization scheme to deal with the
more complex power-law spectrum, and encorporation os processes such as Compton scattering in order
to correctly treat X-ray radiation transfer. In addition, \citep{simmacro2005} modeled
Brackett and Pfund line profiles in YSOs using \py, using the macro atom scheme outlined by Lucy (2002, 2003) applied
to a Hydrogen-only model.

Here, we 
model the UV resonance lines in much the same way as LK02, using the 
fast two-level atom approximation which is appropriate for treating transitions with strong
coupling to the ground state. However, our model has one crucial difference to LK02, as we use
the macro atom scheme in tandem with this method. Hydrogen and Helium are treated
as macro atoms in an attempt to provide a better treatment of optical recombination lines (e.g. the Balmer series, He II 4686, He I lines),
Lyman-$\alpha$, and the recombination continuum which is suspected to play a key role in `filling-in' the Balmer
photoionization edge (see e.g. Knigge \& Drew 1998).

\subsection{Macro-atoms}
In its standard operating mode, \py uses a two-level atom approximation \cite[see e.g.][]{mihalas} for its treatment of lines. This approximation works well when treating so-called `resonance lines' (such as \civ, O\textsc{vi}), in which the excited electron is strongly coupled to the ground state, but breaks down for more complex situations such as recombination cascades. To properly model recombination lines, including the Lyman and Balmer series, a more complete treatment of line transfer is required.

@@

\subsection{Atomic Data}

Atomic data is obtained from the same sources described in LK02 and updated in H13. 
The additional level and line information required to treat Helium as macro atom was 
obtained from \textsc{Topbase} \citep{topbase2005}. We treat Hydrogen as a 20-level atom as in S05, where 
each level is defined by the principal quantum number n, which is appropriate due to
the degeneracy of each level in the Hydrogen atom. Helium II is treated in much the same way,
with 10 levels, but He I has larger energy differentials between different l subshells
and triplet and singlet states. Thus, we still include levels up to n=10 but explicitly 
treat the l-subshells as distinct levels instead of assuming they are perfectly `l-mixed',
which will have the pleasant side effect having the potential to produce triplet lines.

\subsection{Ionization scheme}

Although H13 developed a new ionization scheme which approximates the 
SED in a cell with a broken power law, here we follow LK02 in using the
ionization scheme presented by Mazzali \& Lucy (1993), who specifically
present the formula 
\begin{equation}
\frac{n_{j+1} n_e}{n_j} = W [\xi + W(1-\xi)]
\left(\frac{T_e}{T_R}\right)^{1/2}
\left(\frac{n_{j+1}n_e}{n_j}\right)^*_{T_R}, \label{ionization}
\end{equation}
which, in principle, accounts for ionizations from and
recombinations to all levels of an ion. In this equation, the last
(``starred'') term on the right hand side refers to the Saha
abundances, $W$ is an effective dilution factor, $\xi$ is the
fraction of recombinations going directly to the ground state, and
$T_R$ and $T_e$ are the radiation and electron temperature
respectively. LK02 provide a full description of our implementation.

The reason for using this scheme 
is twofold; First, the SED can be approximated well by a Blackbody due to the 
absence of an X-ray source, and therefore the blackbody treament is appropriate; 
Second, this provides continuity from LK02 and allows us to do a like-for-like comparison,
which is especially useful in assessing the improvement in Hydrogen line emission, and any changes
that might occur in the ionization stages of the resonance line species such as CIV.

@@

\section{Code Validation and Testing}
\subsection{Ionization}

LK02 presented comparisons of between \py and the widely used photoionization code Cloudy, showing good agreement 
of calculated ion fractions between the code. In addition, \cite{tardis} present a code for 
spectral synthesis of Supernova (\textsc{TARDIS}) which compares spectra from a homologous spherical model with
the ML92 ionization scheme. We are therefore confident that our treatment for so-called `simple ions' is correct.

To verify that the macro atom treatment is correct, it is important to compare level population tests 
fro Hydrogen and Helium with that expected from literature. In addition, we check that the dual macro atom/simple ion
implementation still shows good agreement with Cloudy.

\subsection{Non-LTE Level Populations}

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/seaton_emissiv.png}
\caption{Case A comparison between S59 and \py at 10000K and 20000K}
\label{seaton}
\end{figure}

%\begin{figure}
%\includegraphics[width=0.5\textwidth]{figures/caseA.png}
%\caption{Case A comparison between S59 and \py at 10000K and 20000K}
%\end{figure}

%\begin{figure}
%\includegraphics[width=0.5\textwidth]{figures/caseB.png}
%\caption{Case B comparison between S59 and \py at 10000K and 20000K}
%\end{figure}

@@

Figure 3 shows a comparison between the macro atom mode and standard mode of the code using the SV93 model. The ionization state is remarkably similar.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          THE MODEL
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The Benchmark CV Model}
\subsection{Description of Model: Geometry and Kinematics}
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/fig1.eps}
\caption{Cartoon illustrating the three broad classes of sightline discussed in the text.}
\label{cartoon}
\end{figure}

As in LK02, we follow the prescription of SV93 in both our fundamental biconical wind model and velocity law. 
The poloidal velocity, $v_l$, along a streamline in our model is given
by 
\begin{equation}
v_l=v_0+\left[v_{\infty}(r_0)-v_0\right]\frac{\left(l/R_v\right)^{\alpha}}{\left(l/R_v\right)^{\alpha}+1},
\label{v_law}
\end{equation}
where $l$ is distance measured along a poloidal streamline. This
power-law velocity profile was adopted by SV93 in order to give a 
continuous variation in the derivative of the velocity and a 
realistic spread of Doppler-shifted frequencies in the outer portion
of the wind (${l > R_v})$. We have similar requirements.
The initial poloidal velocity of the wind, $v_0$ is (somewhat
arbitrarily) set to $6~\rm{km~s^{-1}}$ for all streamlines, comparable
to the sound speed in the disk photosphere. The
wind then speeds up on a characteristic scale length $R_v$, 
defined as the position along the poloidal streamline at which the
wind reaches half its terminal velocity, $v_{\infty}$.
The terminal poloidal velocity along each streamline is set to a fixed
multiple of the escape velocity at the streamline foot point, 
\begin{equation}
v_{\infty}=fv_{esc},
\label{v_infty}
\end{equation}
so the
innermost streamlines reach the highest velocities.
The power-law index $\alpha$ controls the shape of the velocity
law: as $\alpha$ increases, the acceleration is increasingly
concentrated around $l = R_v$ along each streamline. 
For large $\alpha$, the initial poloidal velocity
stays low near the disk and then increases quickly through $R_v$ to 
values near $v_\infty$.

Table~\ref{wind_param} shows the parameters used in the wind model, and Figure~\ref{cartoon}
shows a sketch of the geometry of our model, with the key parameters labeled.

\begin{table}
\centering
\begin{tabular}{p{3cm}p{4cm}}
\hline Free Parameters 	&	 Value \\ 
\hline \hline 
$M_{WD}$ 	 &	 $0.8 M_{\odot}$ \\ 
$\dot{M}_{acc}$ 	 &	 $3.16\times 10^-9~M_{\odot}yr^{-1}$\\ 
$\dot{M}_{wind}$  &	$3.16\times 10^-8~M_{\odot}yr^{-1}$\\ 
$r_{min}$ 	&	 $4 R_{WD}$\\ 
$r_{max}$ 	&	 $12 R_{WD}$ \\ 
$\theta_{min}$ 	&	 $20.0^{\circ}$ \\ 
$\theta_{max}$ 	&	 $65.0^{\circ}$ \\ 
$\lambda$ 	&	 $0$ \\ 
$v_{\infty}$ 	&	 $3v_{esc}$ \\ 
$R_v$ 	        &	 $7\times10^{11}$cm \\ 
$\alpha$ 	&	 $1.5$ \\
\end{tabular}
\centering
\caption{Wind geometry parameters used in the benchmark CV model.}
\label{wind_param}
\end{table}

\subsection{Disk Treatment}
\py has has some flexibility when treating the accretion disk as a source of photons, in which the disk is broken down into annuli 
such that each annulus contributes an equal amount to the bolometric luminosity. Each annulus can 
be treated either as a blackbody with the corresponding temperature or as a Kurucz stellar atmosphere model (REF)
with the appropriate surface gravity and temperature. Both of these methods have flaws. It is known
from studies of CV accretion disks that they exhibit absorption cores, which can inpact the overall spectrum 
even if a wind is present (REF), so the blackbody treatment is not complete. However, the stellar atmosphere
models also have problems \cite{wadedisk}, and there are clear inconsistencies with using a stellar atmosphere beneath a dense
wind region which itself goes some way to simulating the disk atmosphere. Here we use the Kurucz models,
partly because they provide the most pessimistic prediction due to the strong absorption beyong the Balmer edge, 
which is required to be filled in by the recombination emission in the wind (see Section~\ref{balmerjump}).

\subsubsection{Boundary Layer}
Whether we include one, why why not, referring to He II and ionization.

\subsection{Synthetic Spectra}
Present the spectra.

\begin{figure}
\includegraphics[width=\textwidth]{figures/smplot.pdf}
\caption{Synthetic Spectra computed for sightlines of 22.5, 25, 62.5 and 80 degrees.}
\end{figure}

\subsection{Ionization State}
Present the ionization state with wind plots showing dominant ions.

@@

\section{Discussion of Results}

\subsection{The Balmer Series}

\subsection{UV Resonance Lines}
\subsubsection{OVI and the Auger effect}

\subsection{The Balmer Jump}
\label{balmerjump}

\subsection{Helium Lines}

\subsection{The Line Force}

@@

\section{Conclusions and Future Work}

@@

\subsection{Acknowledgements}
%The work of NH, JHM and CK is supported by the Science and Technology Facilities Council (STFC), via studentships and a consolidated grant, respectively. In particular, we would like to thank Juan Hernandez Santisteban for useful conversations, and for his help extracting SDSS spectra. 
%The authors acknowledge the use of the IRIDIS High Performance Computing Facility, and associated support services at the University of Southampton, in the completion of this work.
